---
title: "a3_task3_james_gately"
author: "James Gately"
date: "10/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(kableExtra)
library(hrbrthemes)

```

## Summary

This markdown file completes the third of four assigned tasks for ESM 206 Assignment 3. In it, I wrangle urchin size distribution data gathered by the SBC LTER program (which can be found [HERE](https://sbclter.msi.ucsb.edu/data/catalog/package/?package=knb-lter-sbc.26)) and do some exploratory data visualization on red and purple urchin test sizes.

### Read in and clean urchin data
```{r}

urchins <- read_csv(here::here("data", "urchins.csv")) %>% 
  clean_names() 

# Note: here::here used due to here() conflict between here::here and lubridate::here

```

### Section 1:

#### Exploratory data visualization
```{r}

# Faceted histrogram
ggplot(data = urchins, aes(x = size)) +
  geom_histogram(bins = 10) +
  facet_wrap(~common_name) +
  labs(x = "test diameter (cm)", y = "count (n)")

```

```{r}

# Boxplot
ggplot(data = urchins, aes(x = common_name, y = size)) +
  geom_boxplot() +
  labs(x = NULL, y = "test diameter (cm)")

```

#### Descriptive statistics
```{r}

# Calculate summary statistics
urchins_stats <- urchins %>% 
  group_by(common_name) %>% 
  summarize(mean(size), 
            median(size), 
            sd(size), 
            n()
            )
# Create table
kable(urchins_stats, 
      col.names = c("Common name", 
                    "Mean test diameter (cm)", 
                    "Median test diameter (cm)", 
                    "Standard deviation (cm)", 
                    "Sample size (n)"),
      caption = "**Table 1.** Summary statistics for Santa Barbara urchin data. ") %>% 
  kable_styling(full_width = FALSE)

```

#### Main takeaways:

We can see that mean test diameters differ between the two species by 3.31 cm -- a percent difference of ~54% -- which encourages further investigation as these two populations may have important differences regarding their test sizes. The boxplot, which shows non-overlapping median diameter sizes, lends additional support to this argument, but further analysis is needed to determine whether differences between the two population's mean test sizes are significant. Interestingly, looking at the histogram, the red urchins appear to be more normally distributed relative to the purple urchins, but both show a relatively normal distribution. Furthermore, the large sample size supports our application of the Central Limit Theorem: we can safely assume a normal distribution for both populations.

### Section 2:

#### Wrangling Purple Urchin data
```{r}

purple_urchins <- urchins %>% 
  filter(common_name == "Purple Urchin") %>%
  mutate(date = mdy(date)) %>% 
  mutate(year = year(date)) %>% 
  mutate(site_full = case_when(
    site == "NAPL" ~ "Naples Reef",
    site == "AQUE" ~ "Arroyo Quemado Reef",
    site == "CARP" ~ "Carpinteria Reef",
    site == "MOHK" ~ "Mohawk Reef",
    site == "IVEE" ~ "Isla Vista Reef"))

purple_urchins_summary <- purple_urchins %>% 
  group_by(site, year) %>% 
  count()

```

#### Exploratory data visualization
```{r}

# Heatmap of purple urchin counts by site
ggplot(data = purple_urchins_summary, aes(y = site, x = year, fill = n)) +
  geom_tile(colour = "white") +
  theme_minimal() +
  labs(x = NULL, y = NULL)

```


